Project Setup

### 2024-11-29 make_readme.qmd
Minor improvements and fixes
Use global language and tidy _quarto.yml
Elaborate on citing code and data

---
title: "Make a README"
lang: en
engine: knitr
---

## What does a README contain?

@@

:   What steps need to be taken to run the project? What software needs to be installed? Mention all dependencies here that are not explicitly managed by `renv`, such as the system dependencies of R packages as well as the version of Quarto. An R package's system dependency is any additional software that you need to install on your computer in order to use a particular R package. For example, the R package [`reticulate`](https://rstudio.github.io/reticulate/) allows to run Python code from within R. However, in order to actually use it, one has to additionally install Python itself as it does not come together with `reticulate` -- rather, it is a system dependency. See @sec-dependencies for additional information.

@@

An overview over the system dependencies of R packages can be created using the function `pak::pkg_sysreqs()`. In combination with `renv`, we can obtain the system dependencies of all R packages in the current project:

```{.r filename="Console"}
pak::pkg_sysreqs(renv::dependencies()$Package)
```

@@

We can see that `make` and `pandoc` were identified as system dependencies. One can obtain their version by running them with the `--version` argument:

@@

>>>>

---
title: "Make a README"
engine: knitr
---

## What Does a README Contain?

@@

:   What steps need to be taken to run the project? What software needs to be installed? R itself and the R packages are already documented as this project uses `renv`. Therefore you can focus on all other dependencies, such as the system dependencies of R packages as well as the version of Quarto.^[As of August 2024, a proposal for `renv` to record the version of Quarto has not been implemented, see [rstudio/renv#1143](https://github.com/rstudio/renv/issues/1143).] Also, don't forget to mention software that you have used for any manual steps. See @sec-dependencies for additional information.

@@

An overview over the system dependencies of R packages can be created using the function `pak::pkg_sysreqs()`. In combination with `renv`, we can obtain the system dependencies of all R packages the current project directly depends on:

```{.r filename="Console"}
# First, install pak
renv::install("pak")

# Then, identify the system dependencies of your direct dependencies
renv::dependencies()$Package |>
  unique() |>
  pak::pkg_sysreqs()
```

@@

We can see that `make` and `pandoc` were identified as system dependencies. Often, one can obtain their version by running them with the `--version` argument:

@@

Of course, all the system dependencies identified until now may have dependencies on their own. Use your own judgement to decide when not to dig deeper.

### 2024-11-29 setup.qmd
Elaborate on citing code and data

## Citing Software and Data

If you rely on software or data by others in your research, the question arises whether and how to cite it in your publications. Put simply, all data relied upon should be cited to allow for precise identification and access. From the "eight core principles of data citation" by @Starr2015, licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/):

@@

When it comes to software, the answer is a little more nuanced – you can consult @fig-software-citation for advice whether to cite it. As with data, citations should allow for exact identification and access. From the six "software citation principles" by @Smith2016, licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/):

@@

To help with generating the citation, you can use the [CiteAs service](https://citeas.org/).

@@

@@

`renv` only records the versions of R packages and of R itself. This means that potential system dependencies of R packages and other tools utilized in the project are not documented anywhere, including Quarto.^[As of August 2024, a proposal to record the version of Quarto has not been implemented, see [rstudio/renv#1143](https://github.com/rstudio/renv/issues/1143).] We will manually write them down when [creating a README](make_readme.qmd). For now, however, there is one simple step you can take to record the version of Quarto (and a few other dependencies). Do run the following:

>>>>

## Cite Data and Software

If you rely on data or software by others in your research, the question arises whether and how to cite it in your publications.

### Data

Put simply, all data relied upon should be cited to allow for precise identification and access. From the "eight core principles of data citation" by @Starr2015, licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/):

@@

Now, add an appropriate citation for the data set to the manuscript. Does your citation adhere to the principles above?

::: {#nte-cite-palmerpenguins .callout-note collapse="true"}
#### Hint for citing the data set

As the data set is from the R package `palmerpenguins`, one can use the function `citation()` to display a suggested citation:

```{r, echo = -1}
invisible(loadNamespace("palmerpenguins")) # Tell renv that we need this package
citation("palmerpenguins")
```

As this can only be run with the package `palmerpenguins` installed, you can also find a [suggested citation on its website](https://allisonhorst.github.io/palmerpenguins/#citation).

Copy the BibTeX entry to the file `Bibliography.bib` and add an identifier between `@Manual{` and the comma, such that the entry's first line reads `@Manual{horst2020,`. Then, add a sentence to the manuscript such as follows:

```md
The analyzed data are by @horst2020.
```

Render the document to check that the citation is displayed properly.

```{.bash filename="Terminal"}
quarto render Manuscript.qmd
```
:::

### Software

When it comes to software, the answer is a little more nuanced due to the large number of involved dependencies. You can consult @fig-software-citation for general advice whether to cite a particular piece of software or not. As with data, citations should allow for exact identification and access. From the six "software citation principles" by @Smith2016, licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/):

@@

In practice, the first step is to identify all pieces of software the project relies on. A few of them are obvious, such as R itself, Quarto, and the $\TeX$ distribution we installed before. Then there are the individual R packages, Quarto extensions, and $\TeX$ packages. All of them, in turn, may have dependencies and it is up to you decide when not to dig deeper. For example, some R packages are only thin wrappers around other R packages or around system dependencies which also might deserve credit. A system dependency is additional software that you require on your computer apart from the R package.

@@

Now, add references for the software you would like to cite to the manuscript. In the following, we will demonstrate this for R and all R packages by using the R package `grateful`. For arbitrary software, you can use the [CiteAs service](https://citeas.org/) to create appropriate citations.

Add the following code chunk to the end of the discussion in the manuscript:

``````{.qmd filename="Manuscript.qmd"}
```{{r}}
#| echo: false

grateful::cite_packages(
    output = "paragraph",
    out.dir = ".",
    omit = NULL,
    dependencies = TRUE,
    passive.voice = TRUE,
    bib.file = "grateful-refs"
)
```
``````

This will automatically create a paragraph citing all used packages and generate the bibliography file `grateful-refs.bib`.^[Note that the detection can fail ] Then, in the YAML header, add `grateful-refs.bib` by setting the `bibliography` as follows:

```{.yml filename="Manuscript.qmd"}
bibliography:
  - Bibliography.bib
  - grateful-refs.bib
```

Use `renv` to view, install, and record the newly used package `grateful`:

```{.r filename="Console"}
renv::status()
renv::install()
renv::snapshot()
```

Finally, render the document again and commit the changes:

```{.bash filename="Terminal"}
quarto render Manuscript.qmd

git status
git add .
git commit -m "Cite data and software"
```

@@

`renv` only records the versions of R packages and of R itself. This means that everything we have not decided to cite in the previous step is not documented anywhere. We will cover system dependencies when [creating a README](make_readme.qmd). For now, however, there is one simple step you can take to record the version of Quarto (and a few other dependencies). Do run the following:
